---
title: "Survival of the Fittest: Cox Regression"
subtitle: "Capstone 6940"
author: "Hazardous Conditions: Kayla Boyd & Hermela Shimelis"
execute:
  echo: true
  warning: false
  message: false
  error: false
format: 
  revealjs:
    scrollable: true
    theme: moon
bibliography: references_2.bib
csl: ieee_2.csl
embed-resources: true
slide-number: false
width: 1920
height: 1080
df-print: paged
html-math-method: katex
editor: visual
pdf-separate-fragments: true
fig-align: center
---

```{r}
#| echo: FALSE
#| message: false
#| warning: false

# Load library
library(dplyr)
library(survival)
library(janitor)
library(magrittr)
library(car)
library(ggplot2)
library(tidyverse)
library(broom)
library(MASS)
library(boot)
library(gtsummary)
library(gt)
library(knitr)
library(kableExtra)
library(broom)
```

# Modeling of Survival after Chemotherapy for Colon Cancer {.text-center}

By Hermela Shimelis

# Introduction

-   The colon cancer data is a built-in data set in the Survival R package [@survival-book, @survival-package]<br>

-   Data set includes 929 subjects with stage B/C colon cancer who were randomized to three treatment groups then followed for 8-years.

    -   Observation, Levamisole (Lev), Levamisole + 5-FU<br>

-   The time to death or progression is given in days. The data set is filtered to evaluate time to death <br><br>

-   ***Objective*****:** Model the relationship between survival time and treatment<br>

# Predictors

```{=html}
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-fymr{border-color:inherit;font-weight:bold;text-align:left;vertical-align:top}
.tg .tg-0pky{border-color:inherit;text-align:left;vertical-align:top}
</style>
```
| Category                | Variables                                       |
|-------------------------|-------------------------------------------------|
| Treatment               | Observation (no treatment)                      |
|                         | Amisole (Lev)                                   |
|                         | Amisole + 5-FU                                  |
| Patient Characteristics | Age                                             |
|                         | Sex                                             |
| Tumor Characteristics   | Colon perforation and obstruction               |
|                         | Adherence to nearby organs                      |
|                         | Tumor differentiation                           |
|                         | Local spread                                    |
|                         | Differentiation of tumor (well, moderate, poor) |
|                         | More than 4 positive lymph nodes                |
| Other                   | Time from surgery to registration               |

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: false

#Load data
colon <- as_tibble(colon)
head(colon)
```

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: false
colon_surv <- colon%>%filter(etype == 2) 

```

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: false
na_counts <- sapply(colon_surv, function(x)sum(is.na(x)))
na_counts

# replace NAs with mode
table(colon_surv$differ)
mode(colon_surv$differ)
median(colon_surv$nodes, na.rm= TRUE)

colon_surv$differ <- if_else(is.na(colon_surv$differ), 2,colon_surv$differ)
colon_surv$nodes <- if_else(is.na(colon_surv$nodes), 2,colon_surv$nodes)
```

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: false
colon_surv <- colon_surv%>%mutate(differentiation = case_when(differ == 1 ~ "well",
                                                              differ == 2 ~ "moderate",
                                                              differ == 3 ~ "poor"),
                                  local_spread = case_when(extent == 1 ~ "submucosa",
                                                           extent == 2 ~ "muscle",
                                                           extent == 3 ~ "serosa",
                                                           extent == 4 ~ "contiguous"),
                                  surg_to_reg_time = case_when(surg == 0~ "short",
                                                               surg == 1 ~ "long"))

```

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: false
library(survival)
library(survminer)
library(ggplot2)
library(gridExtra)
library(dplyr)
```

## Patient characteristics are similar between the three groups {.smaller}

```{=html}
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 2px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 2px;word-break:normal;}
.tg .tg-bobw{font-weight:bold;text-align:center;vertical-align:bottom}
.tg .tg-7zrl{text-align:left;vertical-align:bottom}
.tg .tg-8d8j{text-align:center;vertical-align:bottom}
.tg .tg-j6zm{font-weight:bold;text-align:left;vertical-align:bottom}
</style>
```
|                                             |                            | Observation (%) | Amisole (%)  | Amisole + 5-FU (%) |
|---------------------------------------------|----------------------------|:---------------:|:------------:|:------------------:|
|                                             |                            |      N=315      |    N=310     |       N=304        |
| **Demographics**                            |                            |                 |              |                    |
|                                             | Male                       |   166 (52.3)    |  177 (57.1)  |        141         |
|                                             | Median age (years) \[IQR\] |  60 \[53,68\]   | 61 \[53,69\] |    61 \[52,70\]    |
| **Cancer characteristics**                  |                            |                 |              |                    |
|                                             | Colon obstruction          |    63 (20.0)    |  63 (20.3)   |     54 (17.8)      |
|                                             | Colon perforation          |     9 (2.9)     |   10 (3.2)   |      8 (2.6)       |
|                                             | Adherence to nearby organs |    47 (14.9)    |  49 (15.8)   |     39 (12.8)      |
| Differentiation of tumor                    |                            |                 |              |                    |
|                                             | Well                       |    27 (8.6)     |  37 (11.9)   |      29 (9.5)      |
|                                             | Moderate                   |   236 (74.9)    |  229 (73.9)  |     221 (72.7)     |
|                                             | Poor                       |    52 (16.5)    |  44 (14.2)   |     54 (17.8)      |
| Extent of local spread                      |                            |                 |              |                    |
|                                             | Contiguous                 |    20 (6.3)     |   12 (3.9)   |      11 (3.6)      |
|                                             | Muscle                     |    38 (12.1)    |  36 (11.6)   |     32 (10.5)      |
|                                             | Serosa                     |   249 (79.0)    |  259 (83.5)  |     251 (82.6)     |
|                                             | Submucosa                  |     8 (2.5)     |   3 (1.0)    |      10 (3.3)      |
| More than 4 lymph nodes with cancer         | Yes                        |    87 (27.6)    |  89 (28.7)   |     79 (26.0)      |
| Short time from surgery to registration (%) | Yes                        |    91 (28.9)    |  80 (25.8)   |     76 (25.0)      |

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: False

# Graphical strategies to assess proportionality assumption

# Subset data for modeling
df <- colon_surv%>%dplyr::select(!c(id,study,etype,differ, extent,surg_to_reg_time, nodes))


# sample data for survival plot
set.seed(42)
durations <- rexp(100, 0.1)
event_observed <- rbinom(100, 1, 0.8)
covariate <- sample(0:1, 100, replace = TRUE)

# Kaplan-Meier Survival Distribution S(t)
fit <- survfit(Surv(durations, event_observed) ~ covariate)
plot1 <- ggsurvplot(fit, data = data.frame(durations, event_observed, covariate),
                    xlab = "Survival Time", ylab = "Survival Probability",
                    title = "Kaplan-Meier Survival Distribution S(t)")$plot



# Plot log(-log(S(t))) vs log(Survival Time)
fit1 <- survfit(Surv(time,status) ~ 1, data = df)
surv_prob <- summary(fit1)$surv
log_log_s <- log(-log(surv_prob))
log_time <- log(summary(fit1)$time)

plot2 <- ggplot(data.frame(log_time, log_log_s), aes(x = log_time, y = log_log_s)) +
  geom_line() +
  xlab("log(Survival Time)") +
  ylab("log(-log(S(t)))") +
  ggtitle("log(-log(S(t))) vs log(Survival Time)")

# Schoenfeld Residuals
m1 <- coxph(Surv(time, status) ~ rx, data = df)
schoenfeld_res <- cox.zph(m1)
schoenfeld_df <- as.data.frame(schoenfeld_res$y)


# Check the structure of schoenfeld_df
#str(schoenfeld_df)

# Assuming the residuals are in the first column
plot3 <- ggplot(schoenfeld_df, aes(x = schoenfeld_res$time, y = schoenfeld_df[,1])) +
  geom_point() +
  geom_smooth(method = "loess") +
  xlab("Time") +
  ylab("Schoenfeld Residuals") +
  ggtitle("Schoenfeld Residuals")

# Arrange the plots in a single figure with three panels

grid.arrange(plot1, plot2, plot3, ncol = 2)

```

## Kaplan-Meier Curve Stratified by Treatment Groups

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: False
# Estimate the median survival time among the three groups
survfit(Surv(time,status) ~ rx, data = colon_surv)

# count the number of events after 2080 days, which is the median survival time among the observation group
tt <- colon_surv%>%filter(time > 2083)%>% group_by(rx)%>%summarise(ct = n(),
                                                                   death = sum(status))
# Plot survival curve
fit <- survfit(Surv(time,status) ~ rx, data = colon_surv)
ggsurvplot(fit, data=colon_surv, risk.table = TRUE)

# Estimate the probability of surviving beyond 3000 days
summary(survfit(Surv(time, status) ~ rx, data = colon_surv), times = 3000)

# compare significant difference in survival times between the three groups
survdiff(Surv(time, status)~ rx, data = colon_surv)


```

```{r}
fit <- survfit(Surv(time,status) ~ rx, data = colon_surv)
ggsurvplot(fit, data=colon_surv, risk.table = TRUE)

```

```{r}
#| echo: FALSE
#| message: false
#| warning: false
#| include: false

model <- coxph(formula = Surv(time, status) ~ rx + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)

summary(model)

cox_summary <- tidy(model)

```

## Cox Regression Models

::: panel-tabset
## Model 1: Base Model

```{r}
m0 <- coxph(Surv(time, status) ~ 1, data = df)
summary_m0 = summary(m0)
c_index_m0 <- concordance(m0)


cat("Concordance of the base model:",c_index_m0$concordance)

```

## Model 2: Univariate model

```{r}
Model_2 <- coxph(Surv(time, status) ~ rx, data = df)
```

<br>

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: true

m1 <- coxph(Surv(time, status) ~ rx, data = df)

# Create the regression table and add concordance statistic
summary_table <- tbl_regression(m1, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

# Convert to gt table, increase font size, and adjust width
gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(22),  # Increase font size
    table.align = "center",      # Align table to the left
    table.width = pct(50)     # Make the table wider
  )

# Print the gt table
gt_table
```

## Model 3: All predictors

```{r}
Model_3 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+ local_spread, data = df)
```

<br>

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: True

# Full variables: All predictors
m2 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+ local_spread, data = df)

summary_table <- tbl_regression(m2, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(22),  # Reduce font size
    table.align = "center",    # Align table to the center
    table.width = pct(50),     # Make the table wider
    data_row.padding = px(2)   # Reduce row padding
  )

gt_table
```

## Model 4: Step-wise Selected variables

```{r}
Model_4 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)
```

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: true
# Significant predictors: stepwise-selected variables
m3 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)

summary_table <- tbl_regression(m3, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(22),  # Reduce font size
    table.align = "center",    # Align table to the center
    table.width = pct(50),     # Make the table wider
    data_row.padding = px(2)   # Reduce row padding
  )

gt_table
```

## Proportional Hazard Assumption is Not Met

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: false
cox.zph(m3) # final model with step wise variable selection
zph_test <- cox.zph(m3)
zph_test
# Convert the Schoenfeld residuals test results to a data frame
zph_df <- as.data.frame(zph_test$table)
zph_df$Variable <- rownames(zph_df)

zph_df <- as.data.frame(zph_test$table)
zph_df$Variable <- rownames(zph_df)

zph_df <- zph_df %>%
  mutate(
    chisq = round(chisq, 3),
    p = round(p, 3)
  )
```

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: true
# Print the table in a Quarto presentation slide
zph_df %>%
  kbl(caption = "Schoenfeld Residuals Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, align = "center")
```

```{R}
#| echo: False
#| message: false
#| warning: false
#| include: False

m3 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)

# Schoenfeld Residuals
schoenfeld_res <- cox.zph(m3)
schoenfeld_df <- as.data.frame(schoenfeld_res$y)

# Assuming the residuals are in the first column
plot3 <- ggplot(schoenfeld_df, aes(x = schoenfeld_res$time, y = schoenfeld_df[,1])) +
  geom_point() +
  geom_smooth(method = "loess") +
  xlab("Time") +
  ylab("Schoenfeld Residuals") +
  ggtitle("Schoenfeld Residuals")


```

```{r}
#| echo: false
#| message: false
#| warning: false
#| include: false
# Load necessary libraries
library(survival)
library(ggplot2)
library(gridExtra)

# Fit the Cox proportional hazards model
m3 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + differentiation + node4 + local_spread, data = df)

zph_test <- cox.zph(m3)
zph_test
# Calculate Schoenfeld residuals
schoenfeld_res <- cox.zph(m3)
schoenfeld_df <- as.data.frame(schoenfeld_res$y)
time <- schoenfeld_res$time

# Create a list to store plots
plots <- list()

# Loop through each variable to create plots
for (i in 1:ncol(schoenfeld_df)) {
  p <- ggplot(schoenfeld_df, aes_string(x = "time", y = names(schoenfeld_df)[i])) +
    geom_point() +
    geom_smooth(method = "loess") +
    xlab("Time") +
    ylab("Schoenfeld Residuals") +
    ggtitle(paste("Schoenfeld Residuals for", names(schoenfeld_df)[i]))
  plots[[i]] <- p
}

# Arrange the plots in a grid
grid.arrange(grobs = plots, ncol = 2)
```

## Model 5: Stratified Model

```{r}
Model_5 <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)
```

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: true
# Final Model
m4 <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)

summary_table <- tbl_regression(m4, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(22),  # Reduce font size
    table.align = "center",    # Align table to the center
    table.width = pct(50),     # Make the table wider
    data_row.padding = px(2)   # Reduce row padding
  )
gt_table
```
:::

# Stratified Model Meets Proportional Hazards Assumption

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: false
cox.zph(m4) # final model with step wise variable selection
zph_test <- cox.zph(m4)
zph_test
# Convert the Schoenfeld residuals test results to a data frame
zph_df <- as.data.frame(zph_test$table)
zph_df$Variable <- rownames(zph_df)
zph_df <- zph_df %>%
  mutate(
    chisq = round(chisq, 3),
    p = round(p, 3)
  )


```

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: true
# Print the table in a Quarto presentation slide


zph_df %>%
  kbl(caption = "Schoenfeld Residuals Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, align = "center")

```

# Model Evaluation Metrics <br><br>

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: true
m0 <- coxph(Surv(time, status) ~ 1, data = df)
m1 <- coxph(Surv(time, status) ~ rx, data = df)
m2 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+
              local_spread, data = df)  

m3 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)
m4 <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)

library(knitr)

models <- list(m0, m1, m2, m3, m4)

# Add descriptions for each model
descriptions <- c(
  "Base model",
  "Treatment",
  "Full variables",
  "Stepwise-selected variables",
  "Stratified"
)


# Create a data frame to store results
results <- data.frame(
  Model = character(),
  Description = character(),
  AIC = numeric(),
  BIC = numeric(),
  C_Index = numeric(),
  stringsAsFactors = FALSE
)

# Function to calculate and store metrics for each model
for (i in seq_along(models)) {
  model <- models[[i]]
  
  # Extract AIC and BIC
  aic <- AIC(model)
  bic <- BIC(model)
  
  # Add C-index
  c_index <- concordance(model)$concordance
  
  # Append results to the data frame
  results <- rbind(results, data.frame(
    Model = paste("Model", i),
    Description = descriptions[i],
    AIC = aic,
    BIC = bic,
    C_Index = round(c_index, 3)
  ))
}

# Print the table using kable and kableExtra
results %>%
  kbl(caption = "") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center", font_size = 32 ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:5, width = "10em") %>%
  kable_styling(position = "center")
```

## K-fold Cross Validation

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: False

library(survival)
library(caret)

# Fit the Cox model
cox_model <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 + local_spread, data = df)

# Calculate the original c-index
c_index_original <- survConcordance(Surv(time, status) ~ predict(cox_model), data = df)
#cat("Original c-index:", c_index_original$concordance, "\n")

# Create a function for calculating c-index in each fold using survConcordance
cox_cindex <- function(train_data, test_data) {
  fit <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 + local_spread, data = train_data)
  # Calculate concordance on test data
  c_index <- survConcordance(Surv(test_data$time, test_data$status) ~ predict(fit, newdata = test_data), data = test_data)$concordance
  
  return(c_index)
}

# Perform 5-fold cross-validation with stratification
K <- 5
folds <- createFolds(df$status, k = K, list = TRUE, returnTrain = TRUE)
cv_c_indices <- sapply(folds, function(train_indices) {
  train_data <- df[train_indices, ]
  test_data <- df[-train_indices, ]
  cox_cindex(train_data, test_data) # use the concordance function inside cox_cindex
})


```

```{r}
#| echo: False
#| message: false
#| warning: false
#| include: True
cat("Original c-index:", round(c_index_original$concordance,2), "\n")
cat("Mean cross-validated c-Index:", round(mean(cv_c_indices),2), "\n")

# plot cross-validation c-indices
plot(cv_c_indices, type = "b", xlab = "Fold", ylab = "c-index", main = "c-index across folds")

```

# Conclusions

-   Treatment with Levamisole + 5-FU decreases the hazard of death from colon cancer by 29.5% (HR =0.71, 95% CI: 0.56-0.89; p=0.0035).<br><br>
-   Having more than 4 tumor positive lymph nodes significantly increases the hazard of death by 150.3% (p \<0.0001).<br><br>
-   Having a long wait period from surgery to registration for trial is associated with an increase in the hazard by 99.5% (p=0.01).<br><br>
-   Patients with local tumor spread to the submucosa, muscle and serosa have a reduction in the hazard by 76% (p=.0007), 66% (\<0.001), and 43% (0.004), respectively, compared to those with contiguous organ spread.<br><br>
-   The concordance of the model (0.67) indicates moderate predictive accuracy for survival time.<br><br>
-   Other variables that were not included in the study may contribute to survival time.<br><br>

# Supplementary data

# Predictors

|           |                                                                                   |
|:----------|:----------------------------------------------------------------------------------|
| id:       | id                                                                                |
| study:    | 1 for all patients                                                                |
| rx:       | Treatment - Obs(ervation), Lev(amisole), Lev(amisole)+5-FU                        |
| sex:      | 1=male                                                                            |
| age:      | in years                                                                          |
| obstruct: | obstruction of colon by tumour                                                    |
| perfor:   | perforation of colon                                                              |
| adhere:   | adherence to nearby organs                                                        |
| nodes:    | number of lymph nodes with detectable cancer                                      |
| time:     | days until event or censoring                                                     |
| status:   | censoring status                                                                  |
| differ:   | differentiation of tumour (1=well, 2=moderate, 3=poor)                            |
| extent:   | Extent of local spread (1=submucosa, 2=muscle, 3=serosa, 4=contiguous structures) |
| surg:     | time from surgery to registration (0=short, 1=long)                               |
| node4:    | more than 4 positive lymph nodes                                                  |
| etype:    | event type: 1=recurrence,2=death                                                  |
|           |                                                                                   |

## References

::: {#refs}
:::
